Drought-sensitive trees survive on deep-water access in a tropical forest community

### 2020-08-07 manuscript.rmd
1. Added PET values 2. PLC curves 3. Join A & B goodness of fit curves

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. 
- We tested the role of deep-water access in survival strategies of a diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach. We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We evaluated our ERD estimates with existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that plants mitigate mortality risk from water stress through deep water access.
- The role of deep water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.

Keywords: rooting depths, drought tolerance, mortality, tropical forest, climate change; life-history strategies

@@

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.

In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for 21 species. Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of ~ 320 total). We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD; 1983-2017) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

@@

where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.

We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI--Joseph please add a section). We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011. Exponential form was chosen, because based on AIC criteria it best-fitted all but one species, and for the one species it fitted only slightly worse (Fig. S\@ref(fig:obs-plc-curves); we might want to add specie-specific panel of graphs showing data and fitted curves here), described as

@@

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-plc-curves)).

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$ with blue indicating a smaller value of WGS and red indicating a larger value of WSG.[xx chagne legend to a scale not just two values.]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

## Model for Effective Rooting Depths (ERD)

Water is a key limiting resource for tree growth. Vapour pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_leaf$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (xxadd additive models):

@@

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy}StdVPD_{n} + error
(\#eq:M4)
\end{equation}

We estimated daily value of $K_{leaf,z,n}$ by plugging in $\Psi_{soil,~z,~n}$ in the leaf PLC curves Eq. \@ref(eq:vc-exp) as:

\
\begin{equation}
K_{leaf,~z,~n} = Ae^{-B\Psi_{soil,~z,~n}}
(\#eq:k-soil)
\end{equation}

We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015; see below).

@@

@@

As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) data we collected from the Eddy Flux Tower and VPD (kPa; $\bar{GPP} = m_2VPD^2 + m_1VPD + m_0 + error = StdVPD$; xxAdd SI Fig) and used standardised \bar{GPP} to represent the impact of VPD on growth ($StdVPD$; range 0-1) in Eqs. \@ref(eq:M2)- \@ref(eq:M4). Due to coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the VPD impact may as well be a combined impact of temperature, humidity, and radiation.

To explore the effect of species seasonality in leaf cover on growth, we made an informed assumption of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless). For evergreen species we assumed $LeafCover$ = 1 throughtout the year, whereas for species with all other decisuous classes, we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300), given high forest leaf area index for this period (data not shown). We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). For non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section.

For the combination of each model \@ref(eq:M1)- \@ref(eq:M4), species and hydrological realisation ($sp$), we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$. For each model-species-hydrological realisation we retained only one depth each with the least AIC as the best-fit, and retained only the subset of depths that had a positive Pearson's $\rho$ and *R*^2^ >= 0.5. For each model-species, we defined Effective Rooting Depth (ERD) as the median of this subset of depths and calculated standard error.

@@

## ERD Coordination with Above-ground Hydraulic Traits

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$). See SI for data collection and estimation of these variables (--Brett please add a section).

## Mortality Analyses and Extreme Drought Experience

We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).

@@

### Realised drought risk integral to assessing drought-vulnerability
\
Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, which are commonly  are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$). @Nardini:2015 found that a species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water. Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, (median; `r round(as.numeric(psi.1.7.min$q2.5), 2)`, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.

>>>>

---
title: "Drought-sensitive trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapour pressure deficit, matric potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees' mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
\
Keywords: belowground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies

@@

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information. 
\
In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for 21 species. Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of ~ 320 total). We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(mean(clim$Precip))` mm ($pm$ `r round(sd(clim$Precip))` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain/month). Mean annual Penman-Monteith evapo-transpiration is `r round(mean(clim$pet.PM))` mm ($pm$ `r round(sd(clim$pet.PM))` SD; `r min(clim$Year)`-`r max(clim$Year)`). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

@@

where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively. 
\
We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI--Joseph please add a section). We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011. Exponential form was chosen, because based on AIC criteria it best-fitted all but one species, and for the one species it fitted only slightly worse (Fig. S\@ref(fig:obs-vuln-curves); we might want to add specie-specific panel of graphs showing data and fitted curves here), described as

@@

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-vuln-curves)).

```{r comm-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-vuln-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
```

## Model for Effective Rooting Depths (ERD)
\
Water is a key limiting resource for tree growth. Vapour pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_leaf$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (add additive models xx):

@@

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy}StdVPD_{n} + error
(\#eq:M4)
\end{equation}
\
We estimated daily value of $K_{leaf,z,n}$ by plugging in $\Psi_{soil,~z,~n}$ in the leaf PLC curves Eq. \@ref(eq:vc-exp) as:

\begin{equation}
K_{leaf,~z,~n} = Ae^{-B\Psi_{soil,~z,~n}}
(\#eq:k-soil)
\end{equation}
\
We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015; see below).

@@

\

@@

\
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gCm^2^day^-1^) and VPD we collected from the eddy flux tower (kPa; $\bar{GPP} = -18.8VPD^2 + 14.5VPD + 5.02 + error = StdVPD$; 2012-2017; Fig. S\@ref(fig:gpp-vpd)) and used standardised $\bar{GPP}$ to represent the impact of VPD on growth ($StdVPD$; range 0-1) in Eqs. \@ref(eq:M2) and \@ref(eq:M4). Due to coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the VPD impact may as well be a combined impact of temperature, humidity, and radiation.
\
To explore the effect of species seasonality in leaf cover on growth, we made an informed assumption of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless). For evergreen species we assumed $LeafCover$ = 1 throughtout the year, whereas for species with all other decisuous classes, we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300), given high forest leaf area index for this period (data not shown). We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). For non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardised K within species ($Std.K$; range 0-1) in Eqs. \@ref(eq:M1)- \@ref(eq:M4).
\
For the combination of each model \@ref(eq:M1)- \@ref(eq:M4), species and hydrological realisation, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$. For each model-species-hydrological realisation we retained only one depth each with the least AIC as the best-fit, and retained only the subset of depths that had a positive Pearson's $\rho$ and *R*^2^ >= 0.5. For each model-species, we defined effective rooting depth (ERD) as the median of this subset of depths and calculated standard error.

@@

## ERD Coordination with Above-ground Hydraulic Traits
\
To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$). See SI for data collection and estimation of these variables (--Brett please add a section).

## Mortality Analyses and Extreme Drought Experience
\
We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
\
All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).

@@

### Realised drought risk integral to assessing drought-vulnerability
\
Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, which are commonly  are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$). @Nardini:2015 found that a species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water. Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, (median; `r round(as.numeric(psi.1.7.min$q2.5), 2)`, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (`r round(mean(clim$Precip))` $pm$ `r round(sd(clim$Precip))` mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = `r round(mean(clim$pet.PM))` ($pm$ `r round(sd(clim$pet.PM))` mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study]. We note that the variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
